### 2020-04-04 00-abstract.tex
typos e notazione
qualche typo e aggiunta def per complete chains e freccia doppia,
tolto tau finale da flattened sequences

%Natural computing is an emerging area of research for modeling biological 
%computations by means of computer science formal methods. Reaction systems 
%have been defined in this area as a qualitative abstraction inspired by the 
%functioning of living cells, suitable to model the main
%mechanisms of biochemical reactions.
%Natural computing is an emerging area of research for modeling biological 
%computations by means of computer science formal methods. 
In the area of Natural Computing, reaction systems are
a qualitative abstraction inspired by the 
functioning of living cells, suitable to model the main
mechanisms of biochemical reactions.
This model has already been applied
and extended successfully to various areas of research. Reaction 
systems interact with the environment represented by the context, and
pose problems of implementation, as it is a new computation model. 
In this paper we consider the {\tt link}-calculus, which allows to model 
multiparty interaction in concurrent systems, and show that it allows to
embed reaction systems in a modular way, by representing the behaviour of each entity and
preserving faithfully their features. 
We show the correctness and completeness 
of our embedding.
We illustrate our framework by showing the embedding of 
a few examples expressing computer science
and biological applications.
%how to embed a \emph{lac} operon 
%regulatory network. 
On top of the LTS semantics for reaction systems provided by our framework, 
we then adapt the classical notion (in concurrency) of bisimulation
to make it more suitable for studying properties of reaction systems.
In particular, we define a new assertion language based on regular expressions, 
which allows to specify the properties of interest, and use it to extend Hennessy-Milner logic
to our framework. Interestingly, the novel notion of bisimilarity and logical equivalence that are defined parametrically on some assertion of interest are proved to coincide, like in the classical case.
Finally, our methodology can contribute to increase the expressiveness
of reaction systems, by exploiting the interaction among 
different reaction systems.
%entities and the context.

>>>>

%Natural computing is an emerging area of research for modeling biological 
%computations by means of computer science formal methods. Reaction systems 
%have been defined in this area as a qualitative abstraction inspired by the 
%functioning of living cells, suitable to model the main
%mechanisms of biochemical reactions.
%Natural computing is an emerging area of research for modeling biological 
%computations by means of computer science formal methods. 
In the area of Natural Computing, reaction systems are
a qualitative abstraction inspired by the 
functioning of living cells, suitable to model the main
mechanisms of biochemical reactions.
This model has already been applied
and extended successfully to various areas of research. Reaction 
systems interact with the environment represented by the context, and
pose problems of implementation, as it is a new computation model. 
In this paper we consider a variant of the {\tt link}-calculus, which allows to model 
multiparty interaction in concurrent systems, and show that we can
represent reaction systems as processes 
in a modular way, by representing the behaviour of each entity and
preserving faithfully their features. 
We show the correctness and completeness 
of our embedding.
We illustrate our framework by showing the embedding of 
a few examples expressing computer science
and biological applications.
%how to embed a \emph{lac} operon 
%regulatory network. 
On top of the LTS semantics for reaction systems provided by our framework, 
we then adapt the classical notion (in concurrency) of bisimulation
to make it more suitable for studying properties of reaction systems.
In particular, we define a new assertion language based on regular expressions, 
which allows to specify the properties of interest, and use it to extend Hennessy-Milner logic
to our framework. Interestingly, the novel notion of bisimilarity and logical equivalence that are defined parametrically on some assertion of interest are proved to coincide, like in the classical case.
Finally, our methodology can contribute to increase the expressiveness
of reaction systems, by exploiting the interaction among 
different reaction systems.
%entities and the context.

### 2020-04-04 01-introduction.tex
typos e notazione
qualche typo e aggiunta def per complete chains e freccia doppia,
tolto tau finale da flattened sequences

Then, we present a methodology for verifying formally
properties of reaction systems. The classical notion of bisimulation for 
process algebras allows to consider two processes as equivalent 
where one process can simulate all the actions executed 
by the other one and vice versa. In this paper we define a new
notion of bisimilarity which takes into account the characteristics
of biological systems. We define a new {\em bio-simulation} having in mind
the possibility that two interacting systems may be compared w.r.t.
a subset of the possible biological actions. This is useful for concentrating 
on the sub-model that one may need to study for a specific study or application,
without getting lost in the complexity of the full biological system, or network.
The notion of bio-simulation relies on a new and simple assertion language,
which allows to focus on some properties of the reaction systems to be verified.
In fact, bio-similarity is parametric on a given assertion of interest.
Then we prove that the well-known logical characterisation of bisimilarity in terms of Hennessy-Milner Logic~\cite{HM80} (HML)
can be extended to the case of bio-similarity by tailoring HML formulas to
the same assertion of interest used in bio-similarity.  As HML is a powerful
formalism for specifying properties of labeled transition systems, 
its variant considered here looks a suitable option to specify formulas 
over complex labels by abstracting from unnecessary details.

Our main contributions are as follows:
\begin{itemize}
\item the behaviour of the context, for each single 
entity,
%\footnote{Entities that belong to the `domain' of a RS.}
can be
specified in a recursive way, as ordinary processes;
\item along the same lines, non-deterministic contexts can be seamlessly expressed in our setting; 
\item when deterministic contexts are considered, the c\CNA \
computation is deterministic and mimics the evolution of the corresponding RS;
\item we define an assertion language to specify local properties of 
reaction systems and exploit it to define a novel notion of behavioural equivalence for reaction systems, called bio-similarity;
\item we show that our assertions can be used to extend the Hennessy-Milner logic to our framework, in such a way that logical equivalence of processes coincide with bio-similarity;
\item we sketch how one can express the behaviour of entity mutation,
%change the behaviour of  each single entity $s$  in a
%way that a mutation $s'$ can be expressed; 
in such a way that the mutated entity
$s'$ can  take part to only a subset of rules requiring entity $s$;
\item we show that with a little coding effort, our encoding allows two RSs to
communicate; i.e. we can model scenarios where a subset of those entities that the context  can
provide, are instead provided by a second RS.
\end{itemize}

@@

A preliminary version of this paper appeared in~\cite{BBF19}.
There are several major differences w.r.t. the paper~\cite{BBF19}, which is here extended as follows:
\begin{itemize}
\item here we present several new examples to illustrate our framework,
and give more detailed explanations about its use;
\item we introduce an assertion language to specify the properties of reaction systems;
\item we define the notion of bio-simulation to relate reaction systems and compare their behaviours;
\item we show that our assertion language allows to specify properties extending to our
framework the Hennessy-Milner logic;
\item we include here all proofs of main results.
\end{itemize}

@@

Section \ref{sec:rs} 
describes RSs and their semantics (interactive processes).
Section \ref{sec:ccna} describes briefly the c\CNA \ process algebra and
its operational semantics.
Section \ref{sec:trans} defines the embedding of RSs in
c\CNA \ processes and shows some simple examples to illustrate 
it.
Section~\ref{ex:examples} presents a couple of examples with computer science and biological applications.
%Section \ref{ex:lactose} shows a more complex example taken from
%the literature on RSs and illustrating a \emph{lac} operon.
Section~\ref{sec:biosimulation} presents a methodology for the formal verification of
properties of reaction systems that are expressed in a novel assertion language.
Section~\ref{sec:discussion} presents some features and 
advantages of our embedding for the compositionality of RSs.
Finally, Section~\ref{sec:conclusion}
discusses future work, and concludes.

>>>>

Then, we present a methodology for verifying formally
properties of Reaction Systems. The classical notion of bisimulation for 
process algebras allows to consider two processes as equivalent 
where one process can simulate all the actions executed 
by the other one and vice versa. In this paper we define a new
notion of bisimilarity which takes into account the characteristics
of biological systems. We define a new {\em bio-simulation} having in mind
the possibility that two interacting systems may be compared w.r.t.
a subset of the possible biological actions. This is useful for concentrating 
on the sub-model that one may need to consider for a specific study or application,
without getting lost in the complexity of the full biological system, or network.
The notion of bio-simulation relies on a new and simple assertion language,
which allows to focus on some properties of the Reaction Systems to be verified.
In fact, bio-similarity is parametric on a given assertion of interest.
Then we prove that the well-known logical characterisation of bisimilarity in terms of Hennessy-Milner Logic~\cite{HM80} (HML)
can be extended to the case of bio-similarity by tailoring HML formulas to
the same assertion of interest used in bio-similarity.  As HML is a powerful
formalism for specifying properties of labeled transition systems, 
its variant introduced here looks a suitable option to specify formulas 
over complex labels by abstracting from unnecessary details.

Our main contributions are as follows:
\begin{itemize}
\item the behaviour of the context, for each single 
entity,
%\footnote{Entities that belong to the `domain' of a RS.}
can be
specified in a recursive way, as ordinary processes;
\item along the same lines, non-deterministic contexts can be seamlessly expressed in our setting; 
\item when deterministic contexts are considered, the c\CNA \
computation is deterministic and mimics the evolution of the corresponding RS;
\item we define an assertion language to specify local properties of 
Reaction Systems and exploit it to define a novel notion of behavioural equivalence for Reaction Systems, called bio-similarity;
\item we show that our assertions can be used to extend the Hennessy-Milner logic to our framework, in such a way that logical equivalence of processes coincide with bio-similarity;
\item we sketch how one can express the behaviour of entity mutation,
%change the behaviour of  each single entity $s$  in a
%way that a mutation $s'$ can be expressed; 
in such a way that the mutated entity
$s'$ can  take part to only a subset of rules requiring entity $s$;
\item we show that with a little coding effort, our encoding allows two RSs to
communicate; i.e. we can model scenarios where a subset of those entities that the context  can
provide, are instead provided by a second RS.
\end{itemize}

@@

A preliminary version of this paper appeared in~\cite{BBF19}.
There are several major differences w.r.t. the paper~\cite{BBF19}, which is here extended as follows:
\begin{itemize}
\item here we present several new examples to illustrate our framework,
and give more detailed explanations about its use;
\item we introduce an assertion language to specify the properties of Reaction Systems;
\item we define the notion of bio-simulation to relate Reaction Systems and compare their behaviours;
\item we show that our assertion language allows to specify properties extending to our
framework the Hennessy-Milner logic;
\item we include here all proofs of main results.
\end{itemize}

@@

Section \ref{sec:rs} 
describes RSs and their semantics (interactive processes).
Section \ref{sec:ccna} describes briefly the c\CNA \ process algebra and
its operational semantics.
Section \ref{sec:trans} defines the embedding of RSs in
c\CNA \ processes and shows some simple examples to illustrate 
it.
Section~\ref{ex:examples} presents a couple of examples with computer science and biological applications.
%Section \ref{ex:lactose} shows a more complex example taken from
%the literature on RSs and illustrating a \emph{lac} operon.
Section~\ref{sec:biosimulation} presents a methodology for the formal verification of
properties of Reaction Systems that are expressed in a novel assertion language.
Section~\ref{sec:discussion} presents some features and 
advantages of our embedding for the compositionality of RSs.
Finally, Section~\ref{sec:conclusion}
discusses future work, and concludes.

### 2020-04-04 05-translation.tex
typos e notazione
qualche typo e aggiunta def per complete chains e freccia doppia,
tolto tau finale da flattened sequences

\paragraph{Processes for reactions}
We assume that each reaction $a$ is assigned a progressive number $j$.
The process for reaction $aj=(R_j,I_j,P_j)$ must assert either the possibility to apply the reaction or its impossibility.
The first case happens when all its reactants are present (the link $\link{s_i}{s_o}$ is requested for any $s\in R_j$) and all its inhibitors are absent (the link $\link{\overline{e}_i}{\overline{e}_o}$ is requested for any $e\in I_j$), then the product set is released (the link $\link{\widetilde{c}_i}{\widetilde{c}_o}$ is requested for any $c\in P_j$).
The next case can happen for two reasons: one of the reactants is absent (the link $\link{\overline{s}_i}{\overline{s}_o}$ is requested for some $s\in R_j$) or one of the inhibitors is present (the link $\link{e_i}{e_o}$ is requested for some $e\in I_j$).
The process is recursive so that reactions can be applied at any step.
\[
\begin{array}{lcl@{\quad}l}
P_{aj} \defeq \\ 
 ^{r_j} \backslash  \blockchain{s_i}{s_o}{s \in R_j} \, \link{}{} \blockchain{\overline{e}_i}{\overline{e}_o}{e \in I_j} \, \chainedlink{r_{j+1}}{\noact}\chainedlink{\noact}{p_j}\link{}{} \blockchain{\widetilde{c}_i}{\widetilde{c}_o}{c \in P_j}\, \backslash_{p_{j+1}}.P_{aj} & \mbox{\small  \{$aj$ is applicable\}}\\
 + \ \\
 \sum_{s\in R_j}\startchain{r_j}\chainedlink{\overline{s}_i}{\noact}\chainedlink{\noact}{\overline{s}_o} \chainedlink{r_{j+1}}{\noact}\chainedlink{\noact}{p_{j}}\chainend{p_{j+1}}.P_{aj}  & \mbox{\small  \{$aj$ is not applicable\}}\\
+\\
 \sum_{e\in I_j}\startchain{r_j}\chainedlink{e_i}{\noact} \chainedlink{\noact}{e_o} \chainedlink{r_{i+1}}{\noact}\chainedlink{\noact}{p_{j}}\chainend{p_{j+1}}.P_{aj}  & \mbox{\small \{$aj$ is not applicable\}}\\
\end{array}
\]

@@

In the following use the following conventions for denoting different categories of names:
\begin{itemize}
\item
$\mathit{decs} \defeq \{ s, \overline{s}, \widetilde{s},\widehat{s},\underline{s}~|~ s\in S\}$ is the set of names for decorated entities (without subscripts $_i$ and $_o$);
\item
$\mathit{ents} \defeq \{ d_i, d_o~|~ d\in \mathit{decs}\}$ is the set names for entities;
\item
$\mathit{reacts}\defeq \{r_1,...,r_u\}$ is the set of reaction names associated with each reaction $aj$;
\item 
$\mathit{prods}\defeq \{p_1,...,p_u\}$ is the set of names for production $p_j$, associated with  reaction $aj$.
\end{itemize}

@@

The topmost restriction $\restrict{\mathit{names}}$ appearing in the process $\llbracket{\cal A},\gamma\rrbracket$ serves to guarantee that all names appearing in a link of the chain labelling a transition are matched.
Since all names appearing in any prefix of $\llbracket{\cal A},\gamma\rrbracket$ are restricted, in the transition $\llbracket{\cal A},\gamma\rrbracket \xrightarrow{\restrict{\mathit{names}}\upsilon} P'$ it means that the observation $\restrict{\mathit{names}}\upsilon$ has the form $\link{\silent}{\silent}\dots\link{\silent}{\silent}$, i.e., it is silent, and that $\upsilon$ is solid. Later on we will be interested in reasoning about the actual chain $\upsilon$ used in the transition.  It has the peculiarity to start and end with silent actions and to include all names in $\mathit{reacts}\cup\mathit{prods}\cup\{\mathit{cxt}\}$. As a matter of notation we call such chain $\upsilon$ \emph{complete} (solid and with $\silent$ actions at the extremes). We will then use $\llparenthesis{\cal A},\gamma\rrparenthesis$ to refer to the encoding without topmost name restrictions, i.e.,
$$
\llparenthesis{\cal A},\gamma\rrparenthesis \defeq 
I~|~ \prod_{a\in A} P_a ~|~  \mathit{Cxt}^1~|~\prod_{s\in C_0}P_{s} ~|~ \prod_{s \notin C_0}\overline{P_s} 
$$
and will write $P\xRightarrow{\upsilon} P'$ to mean that $P\xrightarrow{\upsilon} P'$ with $\upsilon$ complete.

@@

Viceversa, it is immediate to define the function $\mathit{unflat}$ to rebuild the complete label from the compact string (here we exploit again the half link and block notation):
$$
\mathit{unflat}(x) \defeq\left\{
\begin{array}{ll}
^{x}_{x} & \mbox{if $x\in \mathit{reacts}\cup\{\mathit{cxt}\}\cup\mathit{prods}$}\\
^{x_i}_{x_i}\backslash^{x_o}_{x_o} & \mbox{if $x\in \mathit{decs}$}\\
\end{array}
\right.
$$
$$
\mathit{unflat}(x_1\dots x_n) \defeq\ ^{\silent}\backslash \mathit{unflat}(x_1) \backslash \dots \backslash \mathit{unflat}(x_1) \backslash_{\silent}
$$

>>>>

\paragraph{Processes for reactions}
We assume that each reaction $a$ is assigned a progressive number $j$.
The process for reaction $aj=(R_j,I_j,P_j)$ must assert either the possibility to apply the reaction or its impossibility.
The first case happens when all its reactants are present (the link $\link{s_i}{s_o}$ is requested for any $s\in R_j$) and all its inhibitors are absent (the link $\link{\overline{e}_i}{\overline{e}_o}$ is requested for any $e\in I_j$), then the product set is released (the link $\link{\widetilde{c}_i}{\widetilde{c}_o}$ is requested for any $c\in P_j$).
The next case can happen for two reasons: one of the reactants is absent (the link $\link{\overline{s}_i}{\overline{s}_o}$ is requested for some $s\in R_j$) or one of the inhibitors is present (the link $\link{e_i}{e_o}$ is requested for some $e\in I_j$).
The process is recursive so that reactions can be applied at any step.
\[
\begin{array}{lcl@{\quad}l}
P_{aj} \defeq \\ 
 ^{r_j} \backslash  \blockchain{s_i}{s_o}{s \in R_j} \link{}{} \blockchain{\overline{e}_i}{\overline{e}_o}{e \in I_j} \chainedlink{r_{j+1}}{\noact}\chainedlink{\noact}{p_j}\link{}{} \blockchain{\widetilde{c}_i}{\widetilde{c}_o}{c \in P_j} \backslash_{p_{j+1}}.P_{aj} & \mbox{\small  \{$aj$ is applicable\}}\\
 + \ \\
 \sum_{s\in R_j}\startchain{r_j}\chainedlink{\overline{s}_i}{\noact}\chainedlink{\noact}{\overline{s}_o} \chainedlink{r_{j+1}}{\noact}\chainedlink{\noact}{p_{j}}\chainend{p_{j+1}}.P_{aj}  & \mbox{\small  \{$aj$ is not applicable\}}\\
+\\
 \sum_{e\in I_j}\startchain{r_j}\chainedlink{e_i}{\noact} \chainedlink{\noact}{e_o} \chainedlink{r_{i+1}}{\noact}\chainedlink{\noact}{p_{j}}\chainend{p_{j+1}}.P_{aj}  & \mbox{\small \{$aj$ is not applicable\}}\\
\end{array}
\]

@@

In the following use the following conventions for denoting different categories of names:
\begin{itemize}
\item
$\mathit{decs} \defeq \{ s, \overline{s}, \widetilde{s},\widehat{s},\underline{s}~|~ s\in S\}$ is the set of names for decorated entities (without subscripts $_i$ and $_o$);
\item
$\mathit{ents} \defeq \{ d_i, d_o~|~ d\in \mathit{decs}\}$ is the set names for entities;
\item
$\mathit{reacts}\defeq \{r_1,...,r_u\}$ is the set of reaction names $r_j$ associated with each reaction $aj$;
\item 
$\mathit{prods}\defeq \{p_1,...,p_u\}$ is the set of names for production $p_j$, associated with  reaction $aj$.
\end{itemize}

@@

The topmost restriction $\restrict{\mathit{names}}$ appearing in the process $\llbracket{\cal A},\gamma\rrbracket$ serves to guarantee that all names appearing in a link of the chain labelling a transition are matched.
Since all names appearing in any prefix of $\llbracket{\cal A},\gamma\rrbracket$ are restricted, in the transition $\llbracket{\cal A},\gamma\rrbracket \xrightarrow{\restrict{\mathit{names}}\upsilon} P'$ it means that the observation $\restrict{\mathit{names}}\upsilon$ has the form $\link{\silent}{\silent}\dots\link{\silent}{\silent}$, i.e., it is silent, and that $\upsilon$ is solid. Later on we will be interested in reasoning about the actual chain $\upsilon$ used in the transition.  It has the peculiarity to start and end with silent actions and to include all names in $\mathit{reacts}\cup\mathit{prods}\cup\{\mathit{cxt}\}$. As a matter of notation we call such chain $\upsilon$ \emph{complete}.

\begin{definition}[Complete Chain]
A chain $\upsilon$ is called \emph{complete} if it is solid (i.e., it contains no virtual link) and has silent actions $\silent$ at its extremes.
We  write $P\xRightarrow{\upsilon} P'$ to mean that $P\xrightarrow{\upsilon} P'$ with $\upsilon$ complete.
\end{definition}

We will then use $\llparenthesis{\cal A},\gamma\rrparenthesis$ to refer to the encoding without topmost name restrictions, i.e.,
$$
\llparenthesis{\cal A},\gamma\rrparenthesis \defeq 
I~|~ \prod_{a\in A} P_a ~|~  \mathit{Cxt}^1~|~\prod_{s\in C_0}P_{s} ~|~ \prod_{s \notin C_0}\overline{P_s} .
$$
and will focus on the complete transitions of $\llparenthesis{\cal A},\gamma\rrparenthesis$.

@@

Viceversa, it is immediate to define the function $\mathit{unflat}$ to rebuild the complete label from the compact string (here we exploit again the half link and block notation):
$$
\mathit{unflat}(x) \defeq\left\{
\begin{array}{ll}
^{x}_{x} & \mbox{if $x\in \mathit{reacts}\cup\{\mathit{cxt}\}\cup\mathit{prods}$}\\
^{x_i}_{x_i}\backslash^{x_o}_{x_o} & \mbox{if $x\in \mathit{decs}$}\\
\end{array}
\right.
$$
$$
\mathit{unflat}(x_1\dots x_n) \defeq\ ^{\silent}\backslash \mathit{unflat}(x_1) \backslash \dots \backslash \mathit{unflat}(x_n) \backslash_{\silent}
$$
